---
title: "chap5"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### The Art of R Programming: A Tour of Statistical Software Design Book by Norman Matloff

## 5.1 Creating Data Frames

```{r}
kids <- c("Jack", "Jill")
ages <- c(12, 10)
d <- data.frame(kids, ages, stringsAsFactors = FALSE)    # by default, TRUE. 
d    # matrix-like viewpoint
```

#### 5.1.1. Accessing Data Frames
d is a list
```{r}
d[[1]]
```

```{r}
d$kids     # clearer and safer than d[[1]] and d[, 1]
```

But we can treat it in a matrix-like fashion as well. For example, we can view column 1
```{r}
d[, 1]
```

```{r}
str(d)
```

#### 5.1.2 Extended Example: Regression Analysis of Exam Grades Continued

```{r}
filepath = file.path(getwd(), 'exams')
examsquiz <- read.table(filepath, header = TRUE)
head(examsquiz)
```

## 5.2 Other Matrix-Lite Operations

#### 5.2.1. Extracting Subdata Frames
```{r}
examsquiz[2:5, ]
```

```{r}
# Since examsquiz[2:5,2] is a vector, R created a vector instead of another data frame
examsquiz[2:5, 2]     
```

```{r}
class(examsquiz[2:5, 2])
```

```{r}
# By specifying drop=FALSE, as described for the matrix case in Section 3.6, we can keep it as a (one- column) data frame.
examsquiz[2:5, 2, drop=FALSE]
```

```{r}
class(examsquiz[2:5, 2, drop=FALSE])
```

Filtering
```{r}
examsquiz[examsquiz$Exam.1 >= 3.8 ,]
```

Same as 
```{r}
subset(examsquiz, Exam.1 <= 3.8)
# e do not need to write this: subset(examsquiz, examsquiz$Exam.1 >= 3.8)
```

#### 5.2.2. More on Treatment of NA Values
In some situations, we need to set the option na.rm=TRUE, explicitly telling R to ignore NA values. 
```{r}
x <- c(2, NA, 4)
mean(x)
```
```{r}
mean(x, na.rm=TRUE)
```

In some cases, we may wish to rid our data frame of any observation that has at least one NA value. A handy function for this purpose is complete.cases().
```{r}
kids <- c("Jack", NA, "Jillian", "John")
states <- c("CA", "MA", "MA", NA)
d4 <- data.frame(kids, states, stringsAsFactors = FALSE)    # by default, TRUE. 
d4
```

```{r}
complete.cases(d4)
```

```{r}
d5 <- d4[complete.cases(d4), ]
d5
```
Cases 2 and 4 were incomplete; hence the FALSE values in the output of complete.cases(d4). We then use that output to select the intact rows.

#### 5.2.3. Using the rbind() and cbind() Functions and Alternatives
```{r}
d
```

In using rbind() to add a row, the added row is typically in the form of another data frame or list.
```{r}
rbind(d, list("Laura", 19))
```

You can also create new columns from old ones
```{r}
eq <- cbind(examsquiz, examsquiz$Exam.2 - examsquiz$Exam.1)
class(eq)
head(eq)
```

```{r}
examsquiz$ExamDiff <- examsquiz$Exam.2 - examsquiz$Exam.1
head(examsquiz)
```

```{r}
examsquiz$ExamDiffSqrd <- (examsquiz$Exam.2 - examsquiz$Exam.1)**2
head(examsquiz)
```


```{r}
examsquiz <- subset(examsquiz, select=-c(ExamDiffSqrd))
head(examsquiz)
```

We can even exploit recycling to add a column that is of a different length than those in the data frame:
```{r}
d
```

```{r}
d$one <- 1
d
```

#### 5.2.4. Applying apply()
```{r}
apply(examsquiz, 1, max)
```

#### 5.2.5. Extended Example: A Salary Study
```{r}
# Setting as.is to TRUE here is simply an alternate way 
# to achieve stringsAsFactors=FALSE
all2006 <- read.csv("2006_partial.csv", header = TRUE, as.is=TRUE)
```

Some filterings...
```{r}
# exclude hourly-wages
all2006 <- all2006[all2006$Wage_Per == "Year", ]

# exclue weird cases
all2006 <- all2006[all2006$Wage_Offered_From > 20000, ]

# exclue hrly prv wg
all2006 <- all2006[all2006$Prevailing_Wage_Amount > 200, ]
```

Create a new column for the ratio between actual wage and prevailing wage:
```{r}
all2006$rat <- all2006$Wage_Offered_From / all2006$Prevailing_Wage_Amount
```

To calculate the median
```{r}
medrat <- function(dataframe) {
  median_rat <- median(dataframe$rat, na.rm = TRUE) 
  return(median_rat)
}
```

Look for those 3 occupations
```{r}
# Use grep() function to identify the rows containing the given job title.
se2006 <- all2006[grep("Software Engineer", all2006), ]
ap2006 <- all2006[grep("Assistant Professor", all2006), ]
cs2006 <- all2006[grep("Computer Scientist", all2006), ]
```

```{r}
print(medrat(se2006))
print(medrat(ap2006))
print(medrat(cs2006))
```

Another aspect of interest was analysis by firm
```{r}
makecorp <- function(corpname) {
  t <- all2006[all2006$Employer_Name == corpname, ]
  return(t)
}
```

```{r}
corplist <- c("MORGAN STANLEY", "ms", 
              "ITT INDUSTRIES", "itt")

for (i in 1:(length(corplist)/2)) {
  fullname <- corplist[2 * i - 1]
  newdtf <- paste(corplist[2 * i], "2006", sep = "")
  assign(newdtf, makecorp(fullname), pos=.GlobalEnv)
}
```

```{r}
ms2006
```

```{r}
itt2006
```

## 5.3 Merging Data Frames
Join
```{r}
kids <- c("Jack", "Jill", "Jillian", "John")
states <- c("CA", "MA", "MA", "MI")
d1 <- data.frame(kids, states, stringsAsFactors = FALSE)    # by default, TRUE. 
d1    # matrix-like viewpoint
```

```{r}
ages <- c(10, 7, 12)
kids <- c("Jill", "Lillian", "Jack")
d2 <- data.frame(ages, kids, stringsAsFactors = FALSE)    # by default, TRUE. 
d2    # matrix-like viewpoint
```

```{r}
d <- merge(d1, d2)
d
```

The merge() function has named arguments by.x and by.y, which handle cases in which variables have similar information but different names in the two data frames.
```{r}
ages <- c(12, 10, 7)
pals <- c("Jack", "Jill", "Lillian")
d3 <- data.frame(ages, pals, stringsAsFactors = FALSE)
d3
```

```{r}
merge(d1, d3, by.x = "kids", by.y = "pals")
```

Duplicate matches will appear in full in the result, possibly in undesir- able ways.
```{r}
d1
```
```{r}
d2a <- rbind(d2, list(15, "Jill"))
d2a
```
There are two Jills in d2a. There is a Jill (age 10) in d1 who lives in Massachu- setts and another Jill (age 15) with unknown residence.

```{r}
merge(d1, d2a)
```
merge(d1, d2a) put "MA" for both Jills (age 10 and age 15) even though we did not specify the residence of another Jill (age 15)

#### 5.3.1. Extended Example: An Employee Database
```{r}
# Could not find the data: DA, DB
```


## 5.4 Applying Functions to Data Frames
#### 5.4.1. Using lapply() and sapply() on Data Frames
```{r}
kids <- c("Jack", "Jill")
ages <- c(12, 10)
d <- data.frame(kids, ages, stringsAsFactors = FALSE)    # by default, TRUE. 
d    # matrix-like viewpoint
```

```{r}
dl <- lapply(d, sort)
dl
```
So, dl is a list consisting of two vectors, the sorted versions of kids and ages.

Note that dl is a list.
```{r}
class(dl)
```

We could coerce it to a data frame.
```{r}
as.data.frame(dl)  # don't do this as it loses its original relation
```
But this would make no sense, as the correspondence between names and ages has been lost. Jack, for instance, is now listed as 10 years old instead of 12.

#### 5.4.2 Extended Example: Applying Logistic Regression Models

Prepare the abalone data:
```{r}
aba <- read.csv("Abalone.data", header = T, as.is = T)
names(aba) <- c("Sex", "Length", "Diameter", "Height", 
                "Whole weight", "Shucked weight", "Viscera weight", 
                "Shell weight", "Rings")
```

We can use sapply() to fit eight single-predictor models—one for each of the eight variables other than gender in this data set—all in just one line of code.
```{r}
# exclude infants from the analysis
abamf <- aba[aba$Sex != "I", ]
```

Fits a logistic model that predicts Sex from that column and hence from that explanatory variable. 
```{r}
lftn <- function(clmn) {
  glm(as.factor(abamf$Sex) ~ clmn, family = binomial)$coef
}
```

Call sapply() on the subdata frame in which column 1, named Sex, has been excluded. In other words, this is an eight-column subframe consisting of our eight explanatory variables. Thus, lftn() is called on each column of that subframe.
```{r}
loall <- sapply(abamf[, -1], lftn)
```

8 pairs of estiamted beta(i).
```{r}
loall
```

```{r}
class(loall)
```

#### 5.4.3. Extended Example: Aids for Learning Chinese Dialects
Could not find the data. 
