---
title: "chap3"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### The Art of R Programming: A Tour of Statistical Software Design Book by Norman Matloff

## 3.1 Creating Matrices
The internal storage of a matrix is in column-major order, meaning that first all of column 1 is stored, then all of column 2, and so on. Using the matrix() function...
```{r}
y <- matrix(1:4, nrow = 2, ncol = 2)    # equivalent to matrix(c(1, 2, 3, 4), nrow = 2)
y
```

```{r}
y[2,]
```

Another way to build y is to specify elements individually:
```{r}
y <- matrix(nrow = 2, ncol = 2)    # Do need to warn R ahead of time that y will be a matrix and give the number of rows and columns
y[1, 1] <- 1
y[2, 1] <- 2
y[1, 2] <- 3
y[2, 2] <- 4
y
```

Set the byrow argument in matrix() to true to indicate that the data is coming in row-major order
```{r}
m <- matrix(1:6, nrow = 2, byrow = T)
m
```

## 3.2 General Matrix Operations
#### 3.2.1 Performing Linear Algebra Operations on Matrices
y
```{r}
y
```

Mathematical matrix multiplication
```{r}
y %*% y    
```

Mathematical multiplication of matrix by scalar
```{r}
3 * y    
```

Mathematical matrix addition
```{r}
y + y   
```

#### 3.2.2 Matrix Indexing
z
```{r}
z <- matrix(c(1, 2, 3, 4, 1, 1, 0, 0, 1, 0, 1, 0), nrow = 4)
z
```

```{r}
z[, 2:3]
```

Extracting rows instead of columns:
```{r}
y <- matrix(c(11, 21, 31, 12, 22, 32), nrow = 3)
y
```

```{r}
y[2:3, ]
```

```{r}
y[2:3, 2]
```

You can also assign values to submatrices:
```{r}
y <- matrix(1:6, nrow = 3)
y
```


```{r}
y[c(1,3), ] <- matrix(c(1, 1, 8, 12), nrow = 2)
y
```

Another example of assignment to submatrices:
```{r}
x <- matrix(nrow = 3, ncol = 3)
y <- matrix(c(4, 5, 2, 3), nrow = 2)
x
y
```

```{r}
x[2:3, 2:3] <- y
x
```

Negative subscripts, used with vectors to exclude certain elements, work the same way with matrices:
```{r}
y <- matrix(1:6, nrow = 3)
y
```
```{r}
y[-2, ]
```

#### 3.2.3 Extended Example: Image Manipulation
Image files are inherently matrices, since the pixels are arranged in rows and columns.
```{r}
library(pixmap)
# mtrush1.pgm downloaded from the following:
# http://heather.cs.ucdavis.edu/~matloff/
mtrush1 <- read.pnm("mtrush1.pgm", cellres = 1)    
# To get rid of Warning message: In rep(cellres, length = 2) : 'x' is NULL so the result will be NULL
# https://stackoverflow.com/questions/20676755/in-r-after-installed-a-package-pixmap-error-in-filefile-open-rb-ca
mtrush1
```

```{r}
plot(mtrush1)
```
What this class consists of:
```{r}
str(mtrush1)
```

The intensities in this class are stored as numbers ranging from 0.0 (black) to 1.0 (white).
```{r}
mtrush1@grey[28, 88]
```

Blot out.
```{r}
mtrush2 <- mtrush1
mtrush2@grey[84:163, 135:177] <- 1
plot(mtrush2)
```

Adding random noise to the picture
```{r}
# adds random noise to img, at the range rows,cols of img; img and the
# return value are both objects of class pixmap; the parameter q
# controls the weight of the noise, with the result being 1-q times the
# original image plus q times the random noise
blurpart <- function(img, rows, cols, q) {
  lrows <- length(rows)
  lcols <- length(cols)
  newimg <- img
  
  # random noise itself is a sample from the uniform distribution U(0,1): 
  # runif(n, min = 0, max = 1)
  randomnoise <- matrix(nrow = lrows, ncol = lcols, runif(lrows * lcols))
  newimg@grey[rows, cols] <- (1 - q) * img@grey[rows, cols] + q * randomnoise  # q controls the weight of the noise, with larger q values producing more blurring.
  return(newimg)
    }
```


```{r}
mtrush3 <- blurpart(mtrush1, 84:163, 135:177, 0.65)
plot(mtrush3)
```

#### 3.2.4 Filtering on Matrices
```{r}
x <- matrix(c(1, 2, 3, 2, 3, 4), nrow = 3)
x
```

```{r}
x[x[, 2] >= 3, ]    # x[,2] >= 3 equals to FALSE TRUE TRUE
```

The filtering criterion can be based on a variable separate from the one to which the filtering will be applied.
```{r}
z <- c(5, 12, 13)
x[z %% 2 == 1, ]
```

```{r}
m <- matrix(1:6, nrow = 3)
m
```

```{r}
m[m[, 1] > 1 & m[, 2] > 5, ]
```
Please note that we used &, the vector Boolean AND operator, rather than the scalar one that we would use in an if statement, &&.


Since matrices are vectors, you can also apply vector operations to them.
```{r}
m <- matrix(c(5, 2, 9, -1, 10, 11), nrow = 3)
m
```

```{r}
which(m > 2)
```
From a vector-indexing point of view, elements 1, 3, 5, and 6 of m are larger than 2

#### 3.2.5 Extended Example: Generating a Covariance Matrix
Suppose that we are working with an n-variate normal distribution. Our matrix will have n rows and n columns, and we wish each of the n variables to have variance 1, with correlation rho between pairs of variables.


```{r}
makecov <- function(rho, n) {
  m <- matrix(nrow = n, ncol = n)
  m <- ifelse(row(m) == col(m), 1, rho)
  return(m)
}
```

For n = 3 and rho = 0.2.
```{r}
makecov(0.2, 3)
```

col() returns the column number of its argument, just as row() does for the row number.
```{r}
row(m)
```

```{r}
col(m)
```

```{r}
row(m) == col(m)
```


## 3.3 Applying Functions to Matrix Rows and Columns

#### 3.3.1 Using the apply() Function 

```{r}
z <- matrix(1:6, nrow = 3)
apply(z, 2, mean)  # For a matrix 1 indicates rows, 2 indicates columns => 2. column apply
```

A function you write yourself is just as legitimate for use in apply() 
```{r}
f <- function(x) x/c(2, 8)
y <- apply(z, 1, f)    # 1. row apply
y
```

You may have been surprised that the size of the result here is 2 by 3 rather than 3 by 2. If the function to be applied returns a vector of k components, then the result of apply() will have k rows.
You can use the matrix transpose func- tion t() to change it if necessary, as follows:
```{r}
t(apply(z, 1, f))
```

For instance, suppose we have a matrix of 1s and 0s and want to create a vector as follows: For each row of the matrix, the corresponding element of the vector will be either 1 or 0, depending on whether the majority of the first d elements in that row is 1 or 0. Here, d will be a parameter that we may wish to vary.
```{r}
copymaj <- function(rw, d) {
  maj <- sum(rw[1:d]) / d
  return(ifelse(maj > 0.5, 1, 0))     # return(if (maj > 0.5) 1 else 0)
}
```

```{r}
x <- matrix(c(1, 1, 1, 0, 
              0, 1, 0, 1, 
              1, 1, 0, 1, 
              1, 1, 1, 1, 
              0, 0, 1, 0), 
            nrow = 4)
x
```

Here, the values 3 and 2 form the actual arguments for the formal argument d in copymaj()
```{r}
apply(x, 1, copymaj, 3)
```

```{r}
apply(x, 1, copymaj, 2)
```

Contrary to common opinion, using apply() will generally not speed up your code. The benefits are that it makes for very compact code, which may be easier to read and modify, and you avoid possible bugs in writing code for looping.

#### 3.3.2 Extended Example: Finding Outliers
Say we have retail sales data in a matrix rs. Each row of data is for a different store, and observations within a row are daily sales figures. As a simple (undoubtedly overly simple) approach, let’s write code to identify the most deviant observation for each store. We’ll define that as the observation fur- thest from the median value for that store.
```{r}
rs <- matrix(c(3, 5, 6, 15, 3, 8, 5, 20, 1), nrow = 3)
rs
```

```{r}
findols <- function(x) {
  # if inner function is short, we can define inside another function
  findol <- function(xrow) {
    mdn <- median(xrow)
    devs <- abs(xrow - mdn)
    return(which.max(devs))    # equivalent to pandas.Series.idxmax
  }
  return(apply(x, 1, findol))
}
```

```{r}
findols(rs)
```


## 3.4 Adding and Deleting Matrix Rows and Columns
Technically, matrices are of fixed length and dimensions, so we cannot add or delete rows or columns. However, matrices can be reassigned.

#### 3.4.1 Changing the Size of a Matrix
```{r}
x <- c(12, 5, 13, 16, 8)
x <- c(x, 20)     # append 20 at very last
x
```

```{r}
x <- c(x[1:3], 20, x[4:6])    # insert 20 in the middle
x
```

```{r}
x <- x[-2:-4]    # delete elements 2 through 4
x
```

```{r}
one <- c(1, 1, 1, 1)
z <- matrix(c(1, 2, 3, 4,
              1, 1, 0, 0,
              1, 0, 1, 0), 
            nrow = 4)
z
```

cbind() creates a new matrix by combining a column of 1s with the columns of z
```{r}
cbind(one, z)
```

You can also use the rbind() and cbind() functions as a quick way to cre-
ate small matrices.
```{r}
q <- cbind(c(1, 2), c(3, 4))
q
```

But be careful. If you are adding rows or columns one at a time within a loop, and the matrix will eventually become large, it’s better to allocate a large matrix in the first place. It will be empty at first, but you fill in the rows or columns one at a time, rather than doing a time-consuming matrix memory alloca- tion each time.

Delete rows or columns by reassignment
```{r}
m <- matrix(1:6, nrow = 3)
m
m <- m[c(1, 3), ]     # select 1st and 3rd row
m
```


#### 3.4.2 Extended Example: Finding the Closest Pair of Vertices in a Graph
Suppose we need a function that inputs a distance matrix, where the element in row i, column j gives the distance between city i and city j and outputs the minimum one-hop distance between cities and the pair of cities that achieves that minimum.
```{r}
# returns the minimum value of d[i, j], i != j, and the row/col attaining
# that minimum, for square symmetric matrix d; no special policy on ties
mind <- function(d) {
  n <- nrow(d)
  
  # add a column to identify row number for apply()
  dd <- cbind(d, 1:n)    # 1:5 is the row number
  
  wmins <- apply(dd[-n, ], 1, imin)
  
  # wmins will be 2xn, 1st row being indices and 2nd being values
  i <- which.min(wmins[2, ])
  j <- wmins[1, i]
  return(c(d[i, j], i, j))
}

# finds the location, value of the minimum in a row x
imin <- function(x) {
  lx <- length(x)
  i <- x[lx]     # original row number
  j <- which.min(x[(i + 1):(lx - 1)])
  k <- i + j
  return(c(k, x[k]))
}
```


```{r}
# q has to be symmetric, because the distance from city i to city j is the same as from j to i
q <- rbind(c(0, 12, 13, 8, 20),
           c(12, 0, 15, 28, 88),
           c(13, 15, 0, 6, 9),
           c(8, 28, 6, 0, 33),
           c(20, 88, 9, 33, 0))
q
```

```{r}
mind(q)
```


If there are multiple cities with the minimum distances. e.g.
```{r}
p <- rbind(c(0, 12, 13, 6, 20),
           c(12, 0, 15, 28, 88),
           c(13, 15, 0, 6, 9),
           c(6, 28, 6, 0, 33),
           c(20, 88, 9, 33, 0))
p
```

```{r}
mind(p)
```


```{r}
minda <- function(d) {
  smallest <- min(d)
  ij <- which(d == smallest, arr.ind = TRUE)    # arr.ind	logical; should array indices be returned when x is an array?
  return(c(smallest, ij))
}
```

## 3.5 More on the Vector/Matrix Distinction
A matrix is just a vector but with two additional attributes: the number of rows and the number of columns.
```{r}
z <- matrix(1:8, nrow = 4)
z
```

As z is still a vector, we can query its length:
```{r}
length(z)
```

But as a matrix, z is a bit more than a vector:
```{r}
class(z)
attributes(z)
```

```{r}
dim(z)
```

```{r}
nrow(z) 
ncol(z)
```

```{r}
nrow
```


## 3.6 Avoiding Unintended Dimension Reduction
```{r}
z <- matrix(1:8, nrow = 4)
r <- z[2, ]
r
```

```{r}
attributes(z)
attributes(r)    # NULL. r is a vector of length 2, rather than a 1-by-2 matrix.
```

```{r}
str(z)
str(r)    # r is a vector of length 2, rather than a 1-by-2 matrix.
```
This seems natural, but in many cases, it will cause trouble in programs that do a lot of matrix operations.

Fortunately, R has a way to suppress this dimension reduction: the drop argument. 
```{r}
r <- z[2, , drop = FALSE]    # to make r a 1-by-2 matrix, not a two-element vector.
r
```

```{r}
dim(r)
str(r)
```

If you have a vector that you wish to be treated as a matrix, you can use the as.matrix() function, as follows:
```{r}
u <- c(1:3)
v <- as.matrix(u)
attributes(u)
attributes(v)
```


## 3.7 Naming Matrix Rows and Columns

```{r}
z <- matrix(1:4, nrow = 2)
colnames(z)
```

```{r}
colnames(z) <- c("a", "b")
z
```
```{r}
z[, "a"]
```


## 3.8 Higher-Dimensional Arrays

```{r}
firsttest <- matrix(c(46, 21, 50, 30, 25, 50), nrow = 3)
firsttest
```


```{r}
secondtest <- matrix(c(46, 41, 50, 43, 35, 50), nrow = 3)
secondtest
```

Now let’s put both tests into one data structure, which we’ll name tests. We’ll arrange it to have two “layers”—one layer per test—with three rows and two columns within each layer. We’ll store firsttest in the first layer and secondtest in the second.

```{r}
tests <- array(data = c(firsttest, secondtest), dim = c(3, 2, 2))
attributes(tests)
```

```{r}
tests
```

```{r}
tests[3, 2, 1]
```
