---
title: "chap7"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### The Art of R Programming: A Tour of Statistical Software Design Book by Norman Matloff

## 7.1 Control Statements
#### 7.1.1 Loops
Python style.
```{r}
x <- c(5, 12, 13)
for (n in x) print(n^2)
```

C-style looping with while and repeat is also available, complete with break, a statement that causes control to leave the loop. 
```{r}
i <- 1
while (i <= 10) i <- i + 4
print(i)
```

```{r}
i <- 1
while(TRUE) {
  i <- i + 4
  if (i > 10) break
}
print(i)
```

The repeat has no Boolean exit condition. Must use break (or return())
```{r}
i <- 1
repeat {
  i <- i + 4
  if (i > 10) break
}
print(i)
```

Another useful statement is next, which instructs the interpreter to skip the remainder of the current iteration of the loop and proceed directly to the next one.
```{r}
# this function is not in the book.
test_next <- function(numbers) {
  for (number in numbers) {
    if (number >= length(numbers)/2) next
    print(number)
  }
}
```

```{r}
test_next(1:5)
```

```{r}
test_next(1:10)
```

The for construct works on any vector, regardless of mode. You can loop
over a vector of filenames, for instance.
```{r}
for (filename in c("file1", "file2")) print(scan(filename))
```

#### 7.1.2 Looping Over Nonvector Sets
R does not directly support iteration over nonvector sets, but there are a couple of indirect yet easy ways to accomplish it:
    • Use lapply(), assuming that the iterations of the loop are independent of each other, thus allowing them to be performed in any order.
    • Use get(). As its name implies, this function takes as an argument a character string representing the name of some object and returns the object of that name. It sounds simple, but get() is a very powerful function.

An example: get()
```{r}
u <- matrix(data = c(1, 2, 3, 1, 2, 4), nrow = 3)
v <- matrix(data = c(8, 12, 20, 15, 10, 2), nrow = 3)

for (m in c("u", "v")) {
  z <- get(m)
  print(lm(z[, 2] ~ z[, 1]))
}
```

#### 7.1.3 If-else
braces are needed in if-else.
The right brace before the else is used by the R parser to deduce that this is an if-else rather than just an if. I
```{r}
test_if_else <- function(r) {
  if (r == 4) {    # this bracket is needed
    r + 1
  } else {
    r + 3
  }
}

y <- test_if_else(4)
print(y)
```

```{r}
r <- 4
y <- if (r == 4) r + 1 else r + 3
y
```

## 7.2 Arithmetic and Boolean Operators and Values

```{r}
x <- c(2, 4, 6)
y <- c(10, 2, 3)
```

Addition, Subtraction, Multiplication, Division
```{r}
x + y
x - y
x * y
x / y
```

Exponentiation
```{r}
x ^ y
```

Modular arithmetic
```{r}
x %% y
10 %% 3
```

Integer division
```{r}
x %/% y
```

Test for equality
```{r}
p <- c(1.0, 2.0)
q <- c(1, 2)
p == q
```

Test for less than or equal to
```{r}
x <= y
```

Test for greater than or equal to
```{r}
x >= y
```

Boolean AND for scalars
```{r}
m <- c(TRUE, FALSE, TRUE)
n <- c(TRUE, TRUE, FALSE)

m[1] && n[1]
```

```{r}
m && n    # looks at just the first elements of each vector
```

```{r}
if (m[1] && n[1]) print('both TRUE')
```

```{r}
if (m && n) print('both TRUE')
```


Boolean OR for scalars
```{r}
m[1] || n[1]
m || n    # looks at just the first elements of each vector
```

Boolean AND for vectors (vector x, y, result)
```{r}
m & n
```

```{r}
if (m & n) print('both TRUE')
```

Boolean OR for vectors (vector x, y, result)
```{r}
m | n
```

Boolean negation
```{r}
!m
```

The Boolean values TRUE and FALSE can be abbreviated as T and F (both must be capitalized). These values change to 1 and 0 in arithmetic expressions:
```{r}
(1 < 2) * (3 < 4)
```

```{r}
(1 < 2) == T
```

```{r}
(1 < 2) == 1
```

## 7.3 Default Values for Arguments

```{r}
default_arg_value_test <- function(x = TRUE) {
  if (x == TRUE) 'x is TRUE' else 'x is FALSE'
}

default_arg_value_test()
```

```{r}
formals(read.table)
```

## 7.4 Return Values
The return value of a function can be any R object. 
You can transmit a value back to the caller by explicitly calling return(). Without this call, the value of the last executed statement will be returned by default.
```{r}
return_value_test <- function(x) {
  if (x < 0) y <- x * (-1) else y <- x
  return(y)
}
```

```{r}
return_value_test(-3)
return_value_test(3)
return_value_test(0)
```

```{r}
return_value_test_no_return_statement <- function(x) {
  if (x < 0) y <- x * (-1) else y <- x
  y
}
```

```{r}
return_value_test_no_return_statement(-3)
return_value_test_no_return_statement(3)
return_value_test_no_return_statement(0)
```

#### 7.4.1 Deciding Whether to Explicitly Call return()
The prevailing R idiom is to avoid explicit calls to return(). One of the reasons cited for this approach is that calling that function lengthens execution time.

#### 7.4.2 Returning Complex Objects
Here is an example of a function being returned:
```{r}
g <- function() {
  t <- function(x) return (x ^ 2)
  return(t)
}
```

```{r}
g()
```

## 7.5 Functions Are Objects
R functions are first-class objects (of the class "function", of course), meaning that they can be used for the most part just like other objects. 
```{r}
g <- function(x) {
  return (x + 1)
}
g(2)
```
Here, function() is a built-in R function whose job is to create functions!

Even the '{' is a function.
```{r}
?'{'
```

```{r}
formals(g)
```

```{r}
body(g)
```

Since functions are objects, you can also assign them, use them as argu- ments to other functions, and so on.
```{r}
f1 <- function(a, b) return (a + b)
f2 <- function(a, b) return (a - b)
f <- f1
f(3, 2)
f <- f2
f(3, 2)
```

```{r}
g <- function(h, a, b) h(a, b)
g(f1, 3, 2)
g(f2, 3, 2)
```

Since functions are objects, you can loop through a list consisting of several functions.
```{r}
g1 <- function(x) return (sin(x))
g2 <- function(x) return(sqrt(x^2 + 1))
g3 <- function(x) return(2 * x - 1)
plot(c(0, 1), c(-1, 1.5)) # prepare the graph, specifying X and Y ranges
for (f in c(g1, g2, g3)) plot(f, 0, 1, add = T) # add plot to existing graph
```

The functions formals() and body() can even be used as replacement functions. You could change the body of a function by assignment:
```{r}
body(g)
body(g) <- quote(2 * x + 3)
body(g)
```
The reason quote() was needed is that technically, the body of a func- tion has the class "call", which is the class produced by quote(). Without the call to quote(), R would try to evaluate the quantity 2 * x + 3.

## 7.6 Environment and Scope Issues
A function—formally referred to as a closure in the R documentation— consists not only of its arguments and body but also of its environment. 

#### 7.6.1 The Top-Level Environment
```{r}
w <- 12
f <- function(y) {
  d <- 8
  h <- function() {
    return(d * (w + y))
  }
  return(h())
}
```

```{r}
environment(f)
```

ls() lists the objects of an environment.
```{r}
ls()
```

More information with ls.str()
```{r}
ls.str()
```

#### 7.6.2 The Scope Hierarchy

#### 7.6.3 More on ls()

#### 7.6.4 Functions Have (Almost) No Side Effects

#### 7.6.5 Extended Example: A Function to Display the Contents of a Call Frame


## 7.7 No Pointers in R


## 7.8 Writing Upstairs

#### 7.8.1 Writing to Nonlocals with the Superassignment Operator
 
#### 7.8.2 Writing to Nonlocals with assign()

#### 7.8.3 Extended Example: Discrete-Event Simulation in R

#### 7.8.4 When Should You Use Global Variables?

####7.8.5 Closures


## 7.9 Recursion

#### 7.9.1 A Quicksort Implementation

#### 7.9.2 Extended Example: A Binary Search Tree


## 7.10 Replacement Functions

#### 7.10.1 What’s Considered a Replacement Function?

#### 7.10.2 Extended Example: A Self-Bookkeeping Vector Class


## 7.11 Tools for Composing Function Code

#### 7.11.1 Text Editors and Integrated Development Environments

#### 7.11.2 The edit() Function


## 7.12 Writing Your Own Binary Operations


## 7.13 Anonymous Functions
