---
title: "Chapter 1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### The Art of R Programming: A Tour of Statistical Software Design Book by Norman Matloff

## 1.1 How to run R
```{r}
rnorm(10)
```

```{r}
mean(rnorm(1000))
sd(rnorm(1000))
```

#### Batch mode
```{r}
pdf("xh.pdf")
hist(rnorm(1000))
dev.off()  # close
```

## 1.2 A first R session
Assign values
```{r}
x <- c(1, 2, 4)
q <- c(x, x, 8)
x[3]  # index starts with 1, not 0
```

What's in the index 0 position?
```{r}
x[0]
```

Slicing
```{r}
x[2:3]
```

Nile dataset
```{r}
mean(Nile)
sd(Nile)
hn <- hist(Nile, breaks = 24)  # change # of bins
```

## 1.3 Introduction to functions
```{r}
oddcount <- function(x){
  k <- 0
  for (n in x) {
    if (n %% 2 == 1) 
      k <- k + 1
  }
  return(k)
}
```
```{r}
oddcount(c(1, 3, 5))
```
```{r}
oddcount(c(0, 1, 2, 3, 7, 9))
```

#### 1.3.1 scope of variable
Global variable
```{r}
f <- function(x) return(x + y)  # warning: no symbol names y in scope
```
```{r}
y <- 3
f(5)
```

#### 1.3.2 Default arguments
```{r}
g <- function(x, y=2, z=3) return(x * y * z)
```
```{r}
g(x = 5)
```

## 1.4 R data structures
#### 1.4.1 Vector
```{r}
x <- 8  # scala is a vector
```

#### 1.4.2 Character strings
```{r}
length(x)
mode(x)
```

```{r}
z <- "abc"
length(z)
mode(z)
```

```{r}
char <- c("abc", "29 88")
length(char)
mode(char)
```

```{r}
u <- paste("abc", "de", "f")
u
```

```{r}
v <- strsplit(u, " ")
v
```

#### 1.4.3 Matrix
```{r}
m <- rbind(c(1, 4), c(2, 3))
m   # index starts with 1, not 0
```

```{r}
c(1, 4) %*% c(2, 3)
```

```{r}
m %*% m
```

```{r}
m[1, ]
```

```{r}
m[, 1]
```

#### 1.4.4 List
```{r}
x <- list(u=2, v="abc")
x$u
x$v
```

```{r}
print(hn)
```

```{r}
str(hn)
```

### 1.4.5 DataFrame
```{r}
mylist = list(kids=c("Jack", "Jill"), ages=c(12, 10))
d <- data.frame(mylist)
d
```

```{r}
d$ages
```

### 1.4.6 Class
```{r}
summary(hn)
```

## 1.5 Extended example
```{r}
examsquiz <- read.table("ExamsQuiz.txt")
class(examsquiz)
```

```{r}
head(examsquiz)
```

```{r}
lma <- lm(examsquiz$V2 ~ examsquiz$V1)
```

```{r}
examsquiz <- read.table("ExamsQuiz.txt", header=TRUE)
head(examsquiz)
```

```{r}
lma <- lm(examsquiz$FINAL ~ examsquiz$MIDTERM)
attributes(lma)
```

```{r}
lma$coefficients
```

```{r}
summary(lma)
```

## 1.6 Startup and shutdown
## 1.7 Getting help
```{r}
?seq  # help(seq)
```
#### 1.7.2 example() function
```{r}
example(seq)
```

```{r}
example(persp)
```

#### 1.7.3 
```{r}
help.search("multivariate normal")
```

